---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggtda

[![Travis-CI Build Status](https://travis-ci.org/rrrlw/ggtda.svg?branch=master)](https://travis-ci.org/rrrlw/ggtda)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/rrrlw/ggtda?branch=master&svg=true)](https://ci.appveyor.com/project/rrrlw/ggtda)
[![Coverage Status](https://img.shields.io/codecov/c/github/rrrlw/ggtda/master.svg)](https://codecov.io/github/rrrlw/ggtda?branch=master)

[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![CRAN version](http://www.r-pkg.org/badges/version/ggtda)](https://CRAN.R-project.org/package=ggtda)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/grand-total/ggtda)](https://CRAN.R-project.org/package=ggtda)

## Overview

The **ggtda** package provides **ggplot2** layers for the visualization of persistence data and other summary data arising from topological data analysis.

## Installation

The development version of **ggtda** can be installed used the **remotes** package:

```r
# install from GitHub
remotes::install_github("rrrlw/ggtda", vignettes = TRUE)
```

For an introduction to **ggtda** functionality, read the vignettes:

```r
# read vignettes
vignette(topic = "intro-ggtda", package = "ggtda")
```

We aim to submit ggtda to [CRAN](https://CRAN.R-project.org) soon.

## Example

**ggtda** visualizes persistence data but also includes stat layers for common TDA constructions. This example illustrates them together. First, generate an artificial "noisy circle" data set and calculate its persistent homology (PH) using the **TDAstats** package, which ports the `ripser` implementation into R:

```{r}
# generate a noisy circle
n <- 36; sd <- .2
set.seed(0)
t <- stats::runif(n = n, min = 0, max = 2*pi)
d <- data.frame(
  x = cos(t) + stats::rnorm(n = n, mean = 0, sd = sd),
  y = sin(t) + stats::rnorm(n = n, mean = 0, sd = sd)
)
# compute the persistent homology
ph <- as.data.frame(TDAstats::calculate_homology(as.matrix(d), dim = 1))
ph <- transform(ph, dimension = as.factor(dimension))
```

Second, pick an example proximity, or diameter, at which to recognize features in the persistence data. This choice corresponds to a specific Čech complex in the filtration underlying the calculation of PH. It is not the best way to identify features (it ignores persistence), but it is easy to link back to the geometric construction.

```{r}
# attach *ggtda*
library(ggtda)
# fix a proximity for a Čech complex
prox <- 2/3
# visualize disks of fixed radii and the Čech complex for this proximity
p_d <- ggplot(d, aes(x = x, y = y)) +
  theme_bw() +
  coord_fixed() +
  stat_disk(radius = prox/2, fill = "aquamarine3", alpha = .15) +
  geom_point()
p_sc <- ggplot(d, aes(x = x, y = y)) +
  theme_bw() +
  coord_fixed() +
  stat_cech2(diameter = prox, fill = "darkgoldenrod", alpha = .1) +
  stat_cech1(diameter = prox, alpha = .25) +
  stat_cech0()
# visualize the persistence data, indicating cutoffs at this proximity
p_bc <- ggplot(ph,
       aes(start = birth, end = death, colour = dimension)) +
  theme_tda() +
  geom_barcode() +
  geom_vline(xintercept = prox, color = "darkgoldenrod", linetype = "dashed")
p_pd <- ggplot(ph,
       aes(start = birth, end = death, colour = dimension)) +
  theme_tda() +
  stat_persistence() +
  lims(x = c(0, NA), y = c(0, NA)) +
  geom_abline(
    intercept = prox, slope = - 1,
    color = "darkgoldenrod", linetype = "dashed"
  )
# combine the plots
gridExtra::grid.arrange(
  p_d, p_sc, p_bc, p_pd,
  layout_matrix = matrix(c(1,3, 2,4), nrow = 2)
)
```

## Contribute

To contribute to **ggtda**, you can create issues for any bugs you find or any suggestions you have on the [issues page](https://github.com/rrrlw/ggtda/issues).

If you have a feature in mind you think will be useful for others, you can also [fork this repository](https://help.github.com/en/articles/fork-a-repo) and [create a pull request](https://help.github.com/en/articles/creating-a-pull-request).
