---
title: "Introduction to the ggtda package"
author: "Raoul R. Wadhwa, J. Cory Brunson, Jacob G. Scott"
date: "13 January 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: barcode1
  title: Persistence barcodes for shapes
  author: 
  - family: Carlsson
    given: Gunnar
  - family: Zomorodian
    given: Afra
  - family: Collins
    given: Anne
  - family: Guibas
    given: Leonidas
  container-title: International Journal of Shape Modeling
  volume: 11
  issue: 2
  page: 149-187
  type: article-journal
  issued: 2005
- id: barcode2
  title: "Barcodes: The persistent topology of data"
  author: 
  - family: Ghrist
    given: Robert
  container-title: Bulletin of the American Mathematical Society
  volume: 45
  issue: 1
  page: 61-75
  type: article-journal
  issued: 2008
- id: flatpersist
  title: A flat persistence diagram for improved visualization of persistent homology
  author: 
  - family: Wadhwa
    given: Raoul
  - family: Dhawan
    given: Andrew
  - family: Williamson
    given: Drew
  - family: Scott
    given: Jacob
  container-title: arXiv
  page: 1812.04567
  type: article-journal
  issued: 2018
- id: diagpersist
  title: Topological persistence and simplification
  author: 
  - family: Edelsbrunner
    given: Herbert
  - family: Letscher
    given: David
  - family: Zomorodian
    given: Afra
  container-title: Discrete and Computational Geometry
  volume: 28
  issue: 4
  page: 511-533
  type: article-journal
  issued: 2002
---

```{r setup-orig, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Purpose

The ggtda package provides users with a ggplot2-compatible tool to visualize persistent homology as part of a topological data analysis pipeline.
This vignette introduces how users can use ggtda to construct publication-quality topological barcodes [@barcode1;@barcode2], flat persistence diagrams [@flatpersist], and diagonal persistence diagrams [@diagpersist].

## Setup and dataset details

We first setup our environment as follows, attaching the necessary R packages (including ggtda itself!) and loading the required dataset.

```{r setup, message = FALSE}
library("ggtda")
library("TDAstats") # to calculate persistent homology

# load dataset
data("annulus2d")
```

The `annulus2d` dataset (visualized below) comes with ggtda, and will be used to demonstrate each form of visualization.

```{r annulus-scatter, echo = FALSE, fig.height = 4, fig.width = 4}
annulus.df <- as.data.frame(annulus2d)
ggplot(annulus.df, aes(x = x, y = y)) +
  geom_point() +
  theme_bw()
```

The `annulus2d` dataset contains 100 points noisily distributed around the circumference of a unit circle.
We will calculate its persistent homology using the TDAstats package (following code block) and then proceed to plot the result with each of the three visualizations using ggtda.

```{r phom}
# calculate persistent homology of annulus and format
annulus.phom <- as.data.frame(calculate_homology(annulus2d))
annulus.phom$dimension <- as.factor(annulus.phom$dimension)
```

We can take a quick look with `head()` and `tail()` to see what the raw form of `annulus2d`'s persistent homology looks like.

```{r peek}
# first few features (0-dimensional)
head(annulus.phom)

# last few features (1-dimensional)
tail(annulus.phom)
```

Clearly, graphical visualizations would permit easier and more effective interpretation than printing raw data.
We will start off by plotting this as a topological barcode.

*N.B.* although TDAstats calculates persistent homology using the Vietoris-Rips complex, persistent homology from another simplicial complex (e.g. Čech complex) can also be visualized with ggtda.
To account for this, the axes will be labeled in a generic manner that should be adapted by users depending on the specific context.

## Topological barcode (aka persistence barcode)

Topological barcodes plot features as vertically stacked horizontal bars.
Here, each bar corresponds to a single feature (single row in `annulus.phom`) with the left boundary of the bar corresponding to the simplicial complex radius at which the feature appears (birth column in `annulus.phom`) and the right boundary of the bar corresponding to the simplicial complex radius at which the feature disappears (death column in `annulus.phom`).
Creating a topological barcode with ggtda is straightforward, allowing users to choose between two equivalent options: `geom_barcode` and `stat_barcode`.

```{r barcode, fig.height = 3, fig.width = 5}
# create topological barcode with geom layer
ggplot(annulus.phom, aes(x = birth, xend = death, colour = dimension)) +
  geom_barcode() +
  xlab("Simplicial complex radius")

# create barcode with stat layer and minimalist theme
ggplot(annulus.phom, aes(xmin = birth, xmax = death, colour = dimension)) +
  stat_barcode() +
  xlab("Simplicial complex radius") +
  theme_barcode()
```

Given that an annulus (or a noisy circle) has a Betti-1 number equal to 1, the dominant 1-dimensional feature (or 1-cycle) is plotted just as expected.
This finding should be replicated in the persistence diagrams as well.

## Flat persistence diagram

## Diagonal persistence diagram

## References