---
title: "Visualizing a simplicial complex with ggtda"
author: "J. Cory Brunson, Raoul R. Wadhwa, Jacob G. Scott"
date: "1 January 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing a simplicial complex with ggtda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Purpose

In addition to visualizing persistent homology with topological barcodes and persistence diagrams, ggtda allows useRs to visualize the underlying simplicial complices for 2-dimensional data.
This feature is particularly valuable in an educational setting, when selecting a radius and plotting a corresponding simplicial complex side-by-side with a topological barcode eases understanding of persistent homology.
This vignette explains how ggtda can be used to quickly and easily plot a simplicial complex.

## Setup and dataset details

We first setup our environment as follows, attaching the necessary R packages (including ggtda itself!) and loading the required datasets.

```{r setup, message = FALSE}
library("ggtda")
library("ggplot2")
library("TDAstats") # to calculate persistent homology

# load dataset
data("unif2d")
data("annulus2d")
```

The `unif2d` dataset (visualized below; left) is included with TDAstats, and a subset of 25 points from it will be used to demonstrate how individual ggtda functions work.
The `annulus2d` dataset (visualized below; right) is included with ggtda, and will be used to demonstrate effective combination use of ggtda functions.

```{r plots, fig.show = 'hold'}
unif_df <- as.data.frame(unif2d[1:25, ])
colnames(unif_df) <- c("x", "y")
annulus_df <- as.data.frame(annulus2d)

# left panel
ggplot(unif_df, aes(x = x, y = y)) +
  geom_point() +
  coord_fixed() +
  theme_bw() +
  xlab("x") + ylab("y")

# right panel
ggplot(annulus_df, aes(x = x, y = y)) +
  geom_point() +
  coord_fixed() +
  theme_bw()
```

## Visualizing a simplicial complex

For the sake of simplicity, we will first demonstrate use of the individual simplicial complex visualization functions with a 25-point subset of `unif2d`.
The [next section](#eg) will use the `annulus2d` dataset to show how the individual functions can work together to create effective visualizations.
We will use a radius of `0.1` for the visualizations in this section.
First, we explore `stat_disk` that, in combination with `geom_point` from ggplot2, will draw proximity circles surrounding each point; since `0.1` is the circle radius, any pair of points less than a distance of `0.2` units away from each other will have overlapping circles, indicating an edge should be drawn between them at this radius.

```{r disk, fig.width = 4, fig.height = 4}
# set radius value
rad_val <- 0.1

# use stat_disk to visualize simplicial complex
ggplot(unif_df, aes(x = x, y = y)) +
  geom_point() +
  stat_disk(radius = rad_val, fill = "blue", alpha = 0.1) +
  coord_fixed() +
  theme_minimal()
```

To directly visualize the edges and better visualize the features themselves, we can use the `stat_vietoris0`, `stat_vietoris1`, and `stat_vietoris2` functions.
`stat_vietoris0` is essentially an alias for `geom_point` and represents the 0-dimensional components (points) of a Vietoris-Rips complex, as follows.
Since the 0-skeleton of a Vietoris-Rips complex is not dependent on a specific radius (`rad_val` above), `stat_vietoris0` requires no parameters.

```{r vietoris0}
ggplot(unif_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris0() +
  theme_minimal()
```

`stat_vietoris1` adds excitement by plotting the 1-dimensional components (edges between point pairs with overlapping circles).
Since these edges are dependent on a specific radius, we provide a value (double `rad_val`) for the `diameter` parameter.
The `alpha` parameter is provided solely for aesthetic purposes.
It is important to note that while `stat_vietoris1` can be used alone (left panel below), it is often most effective when used in conjunction with `stat_vietoris0` (right panel below).

```{r vietoris1, fig.show = 'hold'}
# left panel: stat_vietoris1 alone
ggplot(unif_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris1(diameter = rad_val * 2, alpha = 0.25) +
  theme_minimal()

# right panel: stat_vietoris1 used with stat_vietoris0
ggplot(unif_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris0() +
  stat_vietoris1(diameter = rad_val * 2, alpha = 0.25) +
  theme_minimal()
```

Since `stat_disk` sometimes requires close evaluation to determine if two points are connected, `stat_vietoris1` can be used to add clarity.
The radius is more intuitive to understand `stat_disk`, but topological barcodes and persistence diagrams are generally plotted with the diameter on the relevant axes.
Accordingly, we've parameterized disks using `radius` and complexes using `diameter`.
Note that either parameter can be passed to each stat, and correct values must be provided with care.

```{r vietoris1-disk, fig.height = 5, fig.width = 5}
# combine stat_vietoris1 with stat_disk
ggplot(unif_df, aes(x = x, y = y)) +
  stat_disk(radius = rad_val, fill = "blue", alpha = 0.1) +
  geom_point() +
  stat_vietoris1(diameter = rad_val * 2, alpha = 0.25) +
  theme_minimal()
```

Since edges themselves do not effectively visualize the entire simplicial complex (users do not want to count edges to determine if a set of 7 points are indeed all pairwise connected), `stat_vietoris2` adds the faces of the Vietoris-Rips complex to permit more intuitive interpretation.
`stat_vietoris2` is parameterized similarly to `stat_vietoris1`.
As was the case with `stat_vietoris1`, while `stat_vietoris2` can be used alone (left panel), it is most effectively used when the lower dimensional components of the Vietoris-Rips complex are also plotted (right panel).
Note that the 4-point clique in the top-left corner is shaded slightly darker than the other faces.

```{r vietoris2, fig.show = 'hold', fig.height = 3.45, fig.width = 3.45}
# left panel: stat_vietoris2 alone
ggplot(unif_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris2(diameter = rad_val * 2, alpha = 0.1) +
  theme_minimal()

# right panel: stat_vietoris2 in combination with stat_vietoris1 and stat_vietoris0
ggplot(unif_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris0() +
  stat_vietoris1(diameter = rad_val * 2, alpha = 0.25) +
  stat_vietoris2(diameter = rad_val * 2, alpha = 0.15) +
  theme_minimal()
```

Generally, using `stat_vietoris2` in combination with `stat_disk` creates too much overlap (figure below) and we suggest avoiding use of both in a single panel.

```{r clutter, fig.height = 5, fig.width = 5}
ggplot(unif_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris2(diameter = rad_val * 2, alpha = 0.1) +
  stat_disk(radius = rad_val, fill = "blue", alpha = 0.1) +
  theme_minimal()
```

In addition to the Vietoris-Rips complex, ggtda permits visualization of the Čech complex with the `stat_cech0` (identical to `stat_vietoris0`), `stat_cech1` (identical to `stat_vietoris1`), and `stat_cech2` (distinct from `stat_vietoris2`) functions.
We plot an equilateral triangle using the `stat_vietoris*` functions and `stat_cech*` functions to showcase the difference.
With a diameter of `1.1`, the three points are all pairwise connected, but the three corresponding circles do not all share a single point of overlap (bottom left panel).
Thus, while the Vietoris-Rips complex and the Čech complex share the same 1-skeleton, only the Vietoris-Rips complex has a face between the three points.

```{r cech, fig.height = 3.45, fig.width = 3.45, fig.show = 'hold'}
# create equilateral triangle point cloud
eq_tri <- data.frame(x = c(0, 1, 0.5),
                     y = c(0, 0, sqrt(3)/2))
diam <- 1.1

# top left panel: VR with stat_vietoris*
ggplot(eq_tri, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris2(diameter = diam, fill = "darkgoldenrod", alpha = 0.1) +
  stat_vietoris1(diameter = diam, alpha = 0.25) +
  stat_vietoris0() +
  ggtitle("Vietoris-Rips complex") +
  theme_minimal()

# top right panel: Čech with stat_cech*
ggplot(eq_tri, aes(x = x, y = y)) +
  coord_fixed() +
  stat_cech2(diameter = diam, fill = "darkgoldenrod", alpha = 0.1) +
  stat_cech1(diameter = diam, alpha = 0.25) +
  stat_cech0() +
  ggtitle("Čech complex") +
  theme_minimal()

# bottom left panel: equilateral triangle with stat_disk
ggplot(eq_tri, aes(x = x, y = y)) +
  coord_fixed() +
  stat_disk(radius = diam / 2, fill = "blue", alpha = 0.1) +
  geom_point() +
  theme_minimal()
```

## Persistence, Punctuated {#eg}

The `annulus2d` dataset contains 100 points noisily distributed around the circumference of a unit circle.
We will calculate its persistent homology using the TDAstats package (following code block) and then plot the result with corresponding simplicial complices at various radii.

```{r phom, fig.show = 'hold', fig.height = 3.45, fig.width = 3.45}
# calculate persistent homology of annulus
annulus_phom <- as.data.frame(calculate_homology(annulus2d))
annulus_phom$dimension <- as.factor(annulus_phom$dimension)

# plot topological barcode (left panel)
ggplot(annulus_phom, aes(start = birth, end = death,
                         group = dimension, colour = dimension)) +
  geom_barcode() +
  xlab("Simplicial complex diameter") +
  theme_barcode()

# plot diagonal persistence diagram (right panel)
ggplot(annulus_phom, aes(start = birth, end = death,
                         shape = dimension, colour = dimension)) +
  stat_persistence(diagram = "diagonal") +
  geom_abline(slope = 1, intercept = 0) +
  xlim(c(0, 1.0)) +
  xlab("Feature appearance") +
  ylab("Feature disappearance") +
  coord_fixed() +
  theme_persist()
```

The dominant feature in `annulus2d` is a 1-dimensional loop that we identify in the following code block.

```{r annulus-main}
subset(annulus_phom, dimension == 1)
```

The feature of interest persists the longest and is listed in row 104 of `annulus_phom` (starts at diameter of `0.29`; ends at diameter of `1.44`).
Let's explore simplicial complex visualization prior to the start of the dominant feature (diameter `0.25`), during the presence of the dominant feature (diameter `0.75`), and after the end of the dominant feature (diameter `1.5`).

```{r radii}
diam <- c(0.25, 0.75, 1.5)
rads <- diam / 2
```

We use four panels to properly explore each diameter: (1) topological barcode with a vertical dotted line indicating diameter; (2) persistence diagram with horizontal and vertical lines, with points to the top-left of the intersection indicating current features; (3) simplicial complex visualization with `stat_disk`; and (4) simplicial complex visualization with `stat_vietoris0`, `stat_vietoris1`, and `stat_vietoris2`.
First, we generate the visualization for a diameter of ```r diam[1]```.

```{r before, fig.show = 'hold', fig.width = 3.45, fig.height = 3.45}
# topological barcode
ggplot(annulus_phom, aes(start = birth, end = death,
                         group = dimension, colour = dimension)) +
  geom_barcode() +
  geom_vline(xintercept = diam[1], linetype = "dashed") +
  xlab("Simplicial complex diameter") +
  labs(colour = "dim") +
  theme_barcode() +
  theme(legend.position = c(0.95, 0.5))

# persistence diagram
ggplot(annulus_phom, aes(start = birth, end = death,
                         shape = dimension, colour = dimension)) +
  stat_persistence(diagram = "diagonal") +
  geom_abline(slope = 1, intercept = 0) +
  
  xlim(c(0, 1.5)) + ylim(c(0, 1.5)) +
  xlab("Feature appearance") +
  ylab("Feature disappearance") +
  labs(colour = "dim", shape = "dim") +
  geom_vline(xintercept = diam[1], linetype = "dashed") +
  geom_hline(yintercept = diam[1], linetype = "dashed") +
  coord_fixed() +
  theme_persist() +
  theme(legend.position = c(0.95, 0.4))

# stat_disk
ggplot(annulus_df, aes(x = x, y = y)) +
  geom_point() +
  coord_fixed() +
  stat_disk(radius = rads[1], fill = "blue", alpha = 0.1) +
  theme_minimal()

# stat_vietoris*
ggplot(annulus_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris2(diameter = diam[1], fill = "darkgoldenrod", alpha = 0.1) +
  stat_vietoris1(diameter = diam[1], alpha = 0.25) +
  stat_vietoris0() +
  theme_minimal()
```

Interpretation at diameter ```r diam[1]```:

* topological barcode (top-left panel):
  + the vertical dashed line (current diameter) is to the left of the dominant 1-dimensional (blue) feature, indicating that the dominant feature has not yet started
  + the dashed line does pass through multiple 1-dimensional features of shorter persistence, indicating the presence of at least a couple 1-loops at the current diameter
  + the dashed line also passes through multiple 0-dimensional (red) features; if we consider the 1-skeleton of the Vietoris-Rips complex as a graph, this indicates that the graph is not fully connected and instead is composed of multiple components
* persistence diagram (top-right panel):
  + the dominant feature is in the top-right (not top-left) quadrant, indicating that it has not yet started
  + there are multiple 1-dimensional (blue triangle) features in the top-left quadrant, corresponding to the shorter persistence 1-loops seen in the topological barcode
  + there are multiple 0-dimensional (red circle) features in the top-left quadrant, corresponding to the graph components noted in the topological barcode
* `stat_disk` visualization (bottom-left panel):
  + the dominant feature (1-loop involving most, if not all, of the points) has not yet started because the circumference (loosely) of the annulus is not connected and is instead composed of multiple components
  + there could be 1-loops at this diameter, however, their presence is not clear
* `stat_vietoris*` visualization (bottom-right panel):
  + the dominant feature has not yet started because the 1-skeleton has multiple components
  * there are at least a couple 1-loops at this diameter (adjacent to `(0.0, -1.0)`)

Each of the four panels complement one another by providing overlapping and consistent pieces of information with varying levels of clarity.
Together, we get a good sense of what's going on in `annulus2d` at a diameter of ```r diam[1]```.
Next, we look at diameter ```r diam[2]```.

```{r during, fig.show = 'hold', fig.width = 3.45, fig.height = 3.45}
# topological barcode
ggplot(annulus_phom, aes(start = birth, end = death,
                         group = dimension, colour = dimension)) +
  geom_barcode() +
  geom_vline(xintercept = diam[2], linetype = "dashed") +
  xlab("Simplicial complex diameter") +
  labs(colour = "dim") +
  theme_barcode() +
  theme(legend.position = c(0.95, 0.5))

# persistence diagram
ggplot(annulus_phom, aes(start = birth, end = death,
                         shape = dimension, colour = dimension)) +
  stat_persistence(diagram = "diagonal") +
  geom_abline(slope = 1, intercept = 0) +
  
  xlim(c(0, 1.5)) + ylim(c(0, 1.5)) +
  xlab("Feature appearance") +
  ylab("Feature disappearance") +
  labs(colour = "dim", shape = "dim") +
  geom_vline(xintercept = diam[2], linetype = "dashed") +
  geom_hline(yintercept = diam[2], linetype = "dashed") +
  coord_fixed() +
  theme_persist() +
  theme(legend.position = c(0.95, 0.3))

# stat_disk
ggplot(annulus_df, aes(x = x, y = y)) +
  geom_point() +
  coord_fixed() +
  stat_disk(radius = rads[2], fill = "blue", alpha = 0.05) +
  theme_minimal()

# stat_vietoris*
ggplot(annulus_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris2(diameter = diam[2], fill = "darkgoldenrod", alpha = 0.1) +
  stat_vietoris1(diameter = diam[2], alpha = 0.25) +
  stat_vietoris0() +
  theme_minimal()
```

Interpretation at diameter ```r diam[2]```:

* topological barcode:
  + the dominant feature exists at the current diameter (dashed line intersects with the long, blue bar)
  + no other 1-loop exists at the current diameter (dashed line does not intersect with any other blue bars)
  + only one 0-dimensional feature exists at the current diameter; although the dashed line does not cross any bars representing 0-dimensional (red) features, this is an artifact of the visualization. In fact, once we reach a simplicial complex diameter in which the 1-skeleton is connected (only one component), there is no further point in calculating or visualizing that single feature as the 1-skeleton will not become unconnected at any higher diameters.
* persistence diagram:
  + the dominant feature exists at the current diameter (top-left quadrant in split created by dashed lines)
  + no other 1-loops exist at the current diameter (all other blue triangles are in the lower-left quadrant)
  + only one 0-dimensional feature exists at the current diameter (similar visualization artifact as the topological barcode)
* `stat_disk` visualization:
  + given the widespread overlap, there is clearly a single 0-dimensional feature at this radius
  + there is clearly at least one 1-loop composed of most, if not all, of the point cloud
  + although unlikely based on the plot, it is possible that there exist other 1-loops at this diameter, e.g. at the periphery of the annulus
* `stat_vietoris*` visualization:
  + there is clearly a 0-dimensional feature at this radius (faces connect all vertices)
  + there is clearly at least one 1-loop at this diameter (includes entire annulus)
  + Given the uniform color, it is possible that there is *only* one 1-loop at this diameter.

Lastly, we plot the four panels for diameter ```r diam[3]```.
We leave interpretation of these panels as an exercise to the reader.

```{r after, fig.show = 'hold', fig.width = 3.45, fig.height = 3.45}
# topological barcode
ggplot(annulus_phom, aes(start = birth, end = death,
                         group = dimension, colour = dimension)) +
  geom_barcode() +
  geom_vline(xintercept = diam[3], linetype = "dashed") +
  xlab("Simplicial complex diameter") +
  labs(colour = "dim") +
  theme_barcode() +
  theme(legend.position = c(0.85, 0.5))

# persistence diagram
ggplot(annulus_phom, aes(start = birth, end = death,
                         shape = dimension, colour = dimension)) +
  stat_persistence(diagram = "diagonal") +
  geom_abline(slope = 1, intercept = 0) +
  
  xlab("Feature appearance") +
  ylab("Feature disappearance") +
  labs(colour = "dim", shape = "dim") +
  geom_vline(xintercept = diam[3], linetype = "dashed") +
  geom_hline(yintercept = diam[3], linetype = "dashed") +
  coord_fixed() +
  theme_persist() +
  theme(legend.position = c(0.85, 0.4))

# stat_disk
ggplot(annulus_df, aes(x = x, y = y)) +
  geom_point() +
  coord_fixed() +
  stat_disk(radius = rads[3], fill = "blue", alpha = 0.05) +
  theme_minimal()

# stat_vietoris*
ggplot(annulus_df, aes(x = x, y = y)) +
  coord_fixed() +
  stat_vietoris2(diameter = diam[3], fill = "darkgoldenrod", alpha = 0.1) +
  stat_vietoris1(diameter = diam[3], alpha = 0.25) +
  stat_vietoris0() +
  theme_minimal()
```